---
title: "Exploratory Data Analysis"
output: html_document
---

```{r setup, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(haven)
library(plotly)
library(viridis)
library(patchwork)
library(kableExtra)
library(formattable)
library(data.table)
library(tableone)
```

```{r, include=FALSE, message=FALSE, warning=FALSE}
#Data Import#
nsqipspineCP_1617 = read_csv("./data/nsqipspineCP_1617.csv")
```

```{r, include=FALSE, message=FALSE, warning=FALSE}
#Data cleaning and tidying
cp_spine_tidy = nsqipspineCP_1617 %>%
  mutate_if(is.numeric, ~replace(., . == -99, NA)) %>%
  mutate(
    age_years = age_days/365.25,
    height = height*2.54,
    weight = weight/2.2) %>%
  mutate(
    bmi = weight/((height/100)*(height/100)),
    asa_status = case_when(
      asaclas == "ASA 1 - No Disturb" ~ "1",
      asaclas == "ASA 2 - Mild Disturb" ~ "2",
      asaclas == "ASA 3 - Severe Disturb" ~ "3",
      asaclas == "ASA 4 - Life Threat" ~ "4",
      asaclas == "None assigned" ~ "NA"),
    home_discharge = case_when(
      dischdest == "Expired" ~ "FALSE",
      dischdest == "Facility Which was Home" ~ "TRUE",
      dischdest == "Home" ~ "TRUE",
      dischdest == "Rehab" ~ "FALSE",
      dischdest == "Separate Acute Care" ~ "FALSE",
      dischdest == "Skilled Care, Not Home" ~ "FALSE",
      dischdest == "Unknown" ~ "NA",
      dischdest == "Unskilled Facility Not Home" ~ "FALSE",
      dischdest == "NULL" ~ "NA"),
    level_13 = case_when(
      prncptx == "ARTHRODESIS, ANTERIOR, FOR SPINAL DEFORMITY, WITH OR WITHOUT CAST; 2 TO 3 VERTEBRAL SEGMENTS" ~ "FALSE",
      prncptx == "ARTHRODESIS, ANTERIOR, FOR SPINAL DEFORMITY, WITH OR WITHOUT CAST; 4 TO 7 VERTEBRAL SEGMENTS" ~ "FALSE",
      prncptx == "ARTHRODESIS, ANTERIOR, FOR SPINAL DEFORMITY, WITH OR WITHOUT CAST; 8 OR MORE VERTEBRAL SEGMENTS" ~ "FALSE",
      prncptx == "ARTHRODESIS, POSTERIOR, FOR SPINAL DEFORMITY, WITH OR WITHOUT CAST; UP TO 6 VERTEBRAL SEGMENTS" ~ "FALSE",
      prncptx == "ARTHRODESIS, POSTERIOR, FOR SPINAL DEFORMITY, WITH OR WITHOUT CAST; 7 TO 12 VERTEBRAL SEGMENTS" ~ "FALSE",
      prncptx == "ARTHRODESIS, POSTERIOR, FOR SPINAL DEFORMITY, WITH OR WITHOUT CAST; 13 OR MORE VERTEBRAL SEGMENTS" ~ "TRUE")) %>%
  filter(home_discharge != "NA") %>% 
  select(pufyear_x:ped_spn_post_neurodeftype, age_years, sex, height, weight, bmi, ethnicity_hispanic, race, asa_status, transt, ventilat, asthma, hxcld, oxygen_sup, crf, impcogstat, seizure, nutr_support, hemodisorder, level_13, optime, tothlos, d_opto_dis, death30yn, supinfec, wndinfd, orgspcssi, dehis, oupneumo, pulembol, renainsf, urninfec, cszre, neurodef, cdarrest, othbleed, bleed_ml_tot, othcdiff, othsysep, unplannedreadmission1, reoperation, dischdest, home_discharge)
```

* <a href="#var">Variables</a>
* <a href="#dem">Demographic Information</a>
* <a href="#medhx">Medical History</a>
* <a href="#outcomes">Frequency of Adverse Outcomes</a>

<h1 id="var">Variables</h1>

The NSQIP pediatric spine dataset contains pre-operative, intra-operative, and post-operative variables for over 800 patients with cerebral palsy undergoing spinal fusion surgery. After data cleaning, we selected variables of interest that could help identify risk factors for post-operative discharge to a location other than home, such as to a rehabilitation facility, or to an acute care or skilled nursing care facility. This outcome also included patients who expired in the hospital.

The baseline patient charateristic variables include (but are not limited to): demographic data (age, BMI, race and ethnicity) and medical history (asthma, chronic lung disease, developmental delay, seizures). The intraoperative variables inlcude (but are not limited to): number of spinal levels operated on, operation length of time, bleeding/transfusions. The post-operative outcomes include (but are not limited to) length of stay, death within 30 days, infections, unplanned readmissions, and the outcome we are specifically interested in, location of discharge.

For a full variable list and description, please see the full report. After data cleaning, we had a dataset with 822 observations and 57 variables total.

<h1 id="dem">Demographic Information</h1>

The average age of male and female patients is summarized below:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
cp_spine_tidy %>%
  group_by(sex) %>% 
  summarize(mean_age = mean(age_years), sd = sd(age_years)) %>% 
  rename(Sex = sex, "Average Age (Years)" = mean_age, "Standard Deviation" = sd) %>% 
  knitr::kable(digits = 2) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F,
                position = "left")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cp_spine_tidy %>%
  plot_ly(y = ~age_years, color = ~sex, type = "box") %>% 
  layout(
    xaxis = list(title = " "),
    yaxis = list(title = "Average Age (Years)"))
```

The following plots show the distribution of race and ethnicity in the dataset.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cp_spine_tidy %>%
  count(race) %>% 
  plot_ly(
    x = ~race, y = ~n, type = "bar"
    ) %>% 
  layout(
    xaxis = list(title = " "),
    yaxis = list(title = "Number of Patients"))

cp_spine_tidy %>%
  mutate(ethnicity_hispanic = case_when(
      ethnicity_hispanic == "No" ~ "Non-Hispanic",
      ethnicity_hispanic == "Yes" ~ "Hispanic",
      ethnicity_hispanic == "NULL" ~ "Other/No Answer")) %>% 
  count(ethnicity_hispanic) %>% 
  plot_ly(
    x = ~ethnicity_hispanic, y = ~n, type = "bar"
    ) %>% 
  layout(
    xaxis = list(title = " "),
    yaxis = list(title = "Number of Patients"))
```

<h1 id="medhx">Medical History</h1>

The following table summarizes the frequency of potential pre-operative risk factors in patients' medical histories.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cp_spine_table1 = cp_spine_tidy %>%
mutate(
    sex = factor(sex, ordered = TRUE, levels = c("Female", "Male")),
    race = factor(race, ordered = FALSE, levels = c("American Indian or Alaska Native", "Asian", "Black or African American", "Native Hawaiian or Other Pacific Islander", "Unknown/Not Reported", "White")),
    admit_from = factor(transt, ordered = TRUE, levels = c("Admitted from home/clinic/doctor's office", "Admitted through ER including outside ER with direct hospital admission", "Chronic care/Rehab/Intermediate Care/Spinal Cord", "Transferred from outside hospital (NICU, PICU, Inpatient on General floor, Adult)", "Other")),
    ASAstatus = factor(asa_status, ordered = TRUE, levels = c("1", "2", "3", "4", "5")),
    ventilator_dependence = factor(ventilat, ordered = TRUE, levels = c("No", "Yes")),
    asthma = factor(asthma, ordered = TRUE, levels = c("No", "Yes")),
    home_oxygen = factor(oxygen_sup, ordered = TRUE, levels = c("No", "Yes")),
    cognitive_impairment = factor(impcogstat, ordered = TRUE, levels = c("No", "Yes")),
    seizure_disorder = factor(seizure, ordered = TRUE, levels = c("No", "Yes")),
    nutritional_support = factor(nutr_support, ordered = TRUE, levels = c("No", "Yes")),
    hematologic_disorder = factor(hemodisorder, ordered = TRUE, levels = c("No", "Yes"))
)

myVars <- c("age_years", "sex", "height", "weight", "ASAstatus", "admit_from", "ventilator_dependence", "asthma", "home_oxygen", "cognitive_impairment", "seizure_disorder", "nutritional_support", "hematologic_disorder")
catVars <- c("sex", "ASAstatus", "admit_from", "ventilator_dependence", "asthma", "home_oxygen", "cognitive_impairment", "seizure_disorder", "nutritional_support", "hematologic_disorder")

tab3 <- CreateTableOne(vars = myVars, data = cp_spine_table1, factorVars = catVars)
```

```{r, include=FALSE, message=FALSE, warning=FALSE}
tab3df = print(tab3)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tab3df %>% 
  knitr::kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = F))
```

From this table, we can appreciate that this is not a healthy population. The majority of our patients were American Society of Anesthesiologists (ASA) Physical Status 3-4, indicating that their anesthesiologist felt that they suffered severe systemic disease, with substantive functional limitations. 97% of patients were being cared for at home preoperatively, and 10% were ventilator dependent. 91% were cognitively impaired, with ~50% requiring nutritional support.

From the table below, we see that the majority of patients had neuromuscular-type scoliosis, which results from medical conditions that impair a child's ability to control the muscles that support the spine such as cerebral palsy.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
cp_spine_tidy %>%
  count(ped_spn_class) %>% 
  replace(is.na(.), 0) %>% 
  janitor::clean_names() %>% 
  plot_ly(
    x = ~ped_spn_class, y = ~n, type = "bar"
    ) %>% 
  layout(
    xaxis = list(title = " "),
    yaxis = list(title = "Number of Patients"))
```

<h1 id="outcomes">Frequency of Adverse Outcomes</h1>

Next we set out to determine the frequency of adverse outcomes in our data set.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cp_spine_table2 = cp_spine_tidy %>%
mutate(
    urinary_tract_infection = factor(urninfec, ordered = TRUE, levels = c("No Complication","Urinary Tract Infection")),
    wound_infection = factor(wndinfd, ordered = FALSE, levels = c("No Complication", "Deep Incisional SSI")),
    home_discharge = factor(home_discharge, ordered = TRUE, levels = c("TRUE", "FALSE")),
    reoperation = factor(reoperation, levels = c("No", "Yes")),
    death_in_30_days = factor(death30yn, ordered = TRUE, levels = c("No", "Yes"))
)

myVars <- c("urinary_tract_infection", "wound_infection", "unplanned_readmission", "home_discharge", "reoperation", "death_in_30_days")
catVars <- c("urinary_tract_infection", "wound_infection", "unplanned_readmission", "home_discharge", "reoperation", "death_in_30_days")

tab4 <- CreateTableOne(vars = myVars, data = cp_spine_table2, factorVars = catVars)
```

```{r, include=FALSE, message=FALSE, warning=FALSE}
tab4df = print(tab4)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tab4df %>% 
  knitr::kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = F))
```


Because there are so many different variables and potential relationships to explore, we decided to focus on an outcome variable that was clinically of interest to all three team members working on the project, and that was fairly common (making it better suited to a prediction model). We chose the variable home_discharge, which distinguishes whether or not each patient was able to be discharged to home after surgery. Non-home discharge, on the other hand, included patients who were discharged to rehabilitation or nursing facilities (and who had not previously resided in one), as well as postoperative death.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data_discharge = cp_spine_tidy %>%
  group_by(home_discharge) %>%
  summarize(n = n()) %>% 
  pivot_wider(
    names_from = "home_discharge",
    values_from = "n"
  ) %>% 
  janitor::clean_names()%>% 
  mutate(percent = (false/(true + false))*100)
```
In our dataset, 53 patients (6.45%) had a non-home discharge.

To see our regression analysis on variables that affect home vs. non-home discharge, please see the "Analysis" page or click [here](https://lisaeisler.github.io/p8105_final_website/model.html).